Optimal updating factor in adaptive flat-distribution simulations

### 2016-04-27 doc/invt.tex
invt.tex: small edits

The Wang-Landau (WL) algorithm\cite{wang2001, wang2001pre}
and meta\-dy\-nam\-ics\cite{huber1994, laio2002,
laio2008, barducci2011, sutto2012}
are two widely-used and closely-related\cite{micheletti2004}
techniques for this purpose,
in which a bias potential is gradually built up
during simulation.
%
These techniques employ updating loops
to discourage future visits to configurations
with the current $z$ value
by elevating the bias potential there.
%
The main difference between the two
is that in the WL case,
each updating step is limited to the bin
containing the current $z$,
while metadynamics adopts an extended
Gaussian window
covering several neighboring bins.
%
The latter is more often used in MD simulations
as it guarantees a smooth profile
of the bias potential
suitable for numerical differentiation
in deriving the bias force.
%
However, in both cases, the bias potential
is updated frequently (usually every few MC or MD steps),
which % brings about a side effect of disrupting
disrupts \note{``disturbs''?} the underlying
equilibrium sampling dynamics\cite{laio2002}.
%
Thus, one often has to reduce
the updating magnitude
over the course of simulation
to decrease the error.

The updating schedule,
or the manner of reducing
the updating magnitude,
therefore,
determines the precision of the final bias potential,
hence that of the PMF\cite{liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008, morozov2009,
komura2012, caparica2012, caparica2014,
barducci2008, dickson2011, dama2014}.
%
For the WL algorithm, it is known
that the optimal updating magnitude, $\ln f$,
is given by the product of the
inverse time\cite{liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008}
and the number of bins
(with ``time'' being
the number of simulation steps).
%
The optimal schedule for the metadynamics case,
however, is less studied.

In this study,
we present a method of computing
the optimal schedule
for a general adaptive flat-distribution sampling methods,
including those used in WL and metadynamics simulations.
%
In this method, the optimal schedule is derived
from an equation of motion of a free particle,
with the initial updating magnitude set to
half of the previous equilibrium value.
%
The optimal schedule in the WL case
is shown to be given by
the inverse-time formula,
with the origin of time modified
to satisfy the above optimal initial condition.
%
For metadynamics, however,
the optimal schedule is more complex,
and it can be sensitive to the simulation length
and the width of the Gaussian window.
%
Further, we generalize
the WL updating scheme
to a class of bandpass updating schemes
that are optimal for infinitely long simulations,
and the modified inverse-time schedule is optimal for these schemes.
%
%For finite-length simulations, however,
%metadynamics may deliver better performance.

@@

Here we shall call a simulation that targets
a nearly flat distribution\cite{
dayal2004, trebst2004, zhang2007, barducci2008, singh2011},
$p_i$,
a flat-distribution sampling (FDS) simulation.
%
Roughly speaking, there are two types of FDS methods.
%
The first type,
referred to as the equilibrium FDS, % (E-FDS) methods hereinafter,
uses a fixed bias potential, $V_i$,
(which is estimated beforehand)
during simulation.
%
Examples include the original
entropic or multicanonical sampling\cite{
berg1992, lee1993}
and simulated tempering\cite{
marinari1992, lyubartsev1992}
with $i$ being the index of energy
and that of temperature, respectively.
%
The estimated bias potential, $V_i$,
are often improved
by the histogram, $H_i$, along $i$ accumulated
during simulation.
%
Using $H_i$ as an estimate of $\pi_i$ in
Eq. \eqref{eq:pi_p_phi1}
allows us to correct $V_i$ as
%
\begin{equation}
V^\mathrm{corr}_i
=
V_i
+
\ln \frac{ H_i }
         { p_i }.
\label{eq:vcorr_equil}
\end{equation}
%
\note{This follows from
  $$
  \begin{aligned}
    H_i \approx \pi_i
    &\propto p^*_i \, e^{-V_i},
    \\
    p_i
    &\propto p^*_i \, e^{-V^\mathrm{corr}_i}.
  \end{aligned}
  $$
}
For an unknown system,
it is often difficult
to guess a sufficiently accurate
$V_i$ to start with.
%
Thus,
it can take many iterative rounds of simulations
to achieve the desired flat distribution.
%
\note{Further, a reasonably precise histogram
requires an accumulation period
longer than the autocorrelation time,
which can also be hard to estimate beforehand.}

@@

%For the single-bin scheme
%used in WL algorithm,
Later,
it was found that
a better way
of controlling the schedule $\alpha(t)$
in late stages
is to follow the formula
%
\begin{equation}
  \alpha(t) = \frac{1}{t},
  \label{eq:alpha_invt}
\end{equation}
%
where $t$,
referred to as the ``time'' below,
is the number of steps
from the beginning of the simulation.
%
This simple yet effective formula has attracted
several studies\cite{belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008, morozov2009,
komura2012, caparica2012, caparica2014}.
%
There are, however, some reservation about
the constant of proportionality:
the optimal formula for $\alpha(t)$
might be $c/t$ with a $c$
different from $1.0$\cite{
morozov2007, zhou2008, morozov2009}.

@@

We now fix the endpoint value $q(T)$
and variate the curve, $q(t)$ ($0 < t < T$).
%
This is equivalent to variating $u$
with fixed endpoints,
$u\bigl( q(0) \bigr)  = e^{- q(T)}$
and
$u\bigl( q(T) \bigr) = 1$.
%
Note also that the first term on the right-hand side
of Eq. \eqref{eq:x2t_average} is fixed during
the process.
%
%We shall further demand that
%%
%\begin{equation}
%  \lim_{t \to \infty} q(t) \to \infty.
%  \label{eq:qt_limit}
%\end{equation}
%%
%Then, for a large $t$,
%the second term on the right-hand side
%of Eq. \eqref{eq:xt_solution} can be neglected, and
%%
%\begin{align}
%\left\langle x^2(t) \right\rangle
%%&=
%%\int_0^t \int_0^t \dot u(t') \, \dot u(t'')
%%    \left\langle \xi(t') \xi(t'') \right\rangle dt'' \, dt'
%%\notag
%%\\
%&=
%\int_0^t \int_0^t
%  \dot u\bigl( q(t') \bigr) \,
%  \dot u\bigl( q(t'') \bigr) \,
%  \kappa(t' - t'') \, dt'' \, dt'.
%\label{eq:x2t_average_asym}
%\end{align}
%
By the Euler-Lagrange equation, we get
$$
\begin{aligned}
0
&=
\frac{d}{dt} \int_0^T
  \dot u\bigl( q(t') \bigr) \, \kappa(t' - t) \, dt'
\\
&= \int_0^T
  \ddot u\bigl( q(t') \bigr) \, \kappa(t' - t) \, dt',
\end{aligned}
$$
where we have dropped the boundary terms
by assuming Eq. \eqref{eq:kappa_limit}
and a large $T$.
%
%In fact, we can repeat the process $n$ times, and
%$$
%\int_0^t u^{(n)}(t') \, \kappa(t' - \tau) \, dt' = 0,
%\qquad (n \ge 1)
%$$
%
In order for this to hold for any $t$,
we must have
%
$
\ddot u\bigl( q(t') \bigr) = 0
,
$
or
$
\dot u\bigl( q(t') \bigr) = c
,
$
%
where $c$ is a constant of $t'$.
%
Using Eq. \eqref{eq:u_def},
we get
$$
e^{ q(t') - q(T) }
=
c \, (t' + t_0),
$$
where $t_0$ is a constant.
%
Taking the logarithm,
and differentiating both sides
with respect to $t'$
yields Eq. \eqref{eq:alpha_invt1}.

@@

%It is worth pointing out
Finally, there is a stability condition.
%
In a \emph{stable} updating scheme,
no eigenmode grows indefinitely,
which means all eigenvalues $\lambda_k$'s
are nonnegative.

@@

In the asymptotic regime
(i.e., if the simulation length, $T$, is sufficiently long),
we can approximate $\kappa_k(t)$ as
%
\begin{equation}
  \kappa_k(t) \approx \Gamma_k \, \delta(t)
  ,
\label{eq:kappa_delta}
\end{equation}
%
where
\begin{equation}
  \Gamma_k = \int_{-\infty}^\infty \kappa_k(t) \, dt,
  \notag
  %\label{eq:Gamma_integral}
\end{equation}
or in discrete time,
\begin{equation}
  \Gamma_k = \sum_{t = -\infty}^\infty \kappa_k(t).
  \label{eq:Gamma_sum}
\end{equation}
%
Then, Eq. \eqref{eq:error_asym} is simplified as
%
\begin{align}
  \Err_A
  &=
  \int_0^T
  \sum_{k = 2}^n
  \Gamma_k \, \dot u_k^2\bigl( q(t) \bigr) \, dt.
\label{eq:error_asym1}
\end{align}
%
In this way,
the underlying sampling process
affects the error only through the set of numbers,
$\Gamma_k$'s.
%
In Appendix \ref{sec:Gamma},
we evaluate $\Gamma_k$'s in two special cases,
and show how to measure them
from an adaptive FDS simulation.

@@

We shall use this result to model the initial error
before starting the schedule $\alpha(t)$ to be optimized.
%
That is, we assume
$\alpha(t) = \alpha_0$
for $t < 0$, and
the system has entered the equilibrium
under a fixed $\alpha_0$ at $t = 0$.
%
We can then use Eq. \eqref{eq:y2_eql}
for the $\langle y_k^2(0) \rangle$'s in Eq. \eqref{eq:error_res}.
%
Then,
the residual error at the end of period $T$
is given by
%
\begin{equation}
  \Err_R
  =
  \sum_{k = 2}^n
      \frac 1 2 \, \alpha_0 \, \Gamma_k \, \lambda_k \,
      e^{  - 2 \, \lambda_k \, q(T) }
  .
  \label{eq:error_res1}
\end{equation}
%
Note that the residual error
depends on $\alpha(t)$ only through
the integral $q(T) = \int_0^T \alpha(t) \, dt$.
%
Thus, if we variate $\alpha(t)$ under a fixed $q(T)$,
this residual part can be ignored,
and the optimization can be done exclusively
on the asymptotic part, $\Err_A$.

@@

As shown above, with a fixed value of $q(T)$,
the optimization of the total error is equivalent to
that of the asymptotic part.
%
We can rewrite Eq. \eqref{eq:error_asym1} as an action
%
\begin{equation}
  \Err_A
  =
  \int_0^T
    {\mathcal L} \bigl[ q(t)\bigr]
    \, dt
  ,
%\notag
\label{eq:error_asym_Lagrangian}
\end{equation}
%
where the Lagrangian,
%
\begin{align}
  {\mathcal L} \bigl[ q(t) \bigr]
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \dot u_k^2\bigl[ q(t) \bigr]
  \notag
  \\
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \lambda_k^2 \, u_k^2 \bigl[ q(t) \bigr]
  \; \dot q^2( t ),
\notag
\end{align}
%
represents a free particle
with a position-dependent mass.
%
Since the Lagrangian
does not explicitly depend on time, $t$,
the Hamiltonian
%
\begin{equation}
  \mathcal H
  =
  \frac{ \partial \mathcal L }
       { \partial \dot q     }
  \, \dot q
  -
  \mathcal L
  =
  2 \, \mathcal L
  - \mathcal L
  =
  \mathcal L,
  %=
  %\mathrm{const.}
  \notag
  %\label{eq:error_asym_Hamiltonian}
\end{equation}
%
which coincides with the Lagrangian,
is conserved.
%
So we may set
%
\begin{equation}
  \mathcal L
  =
    \sum_{ k = 2 }^n
      \Gamma_k \, \lambda_k^2
      \, u_k^2 \bigl[ q(t) \bigr]
  \;
  \dot q^2(t)
  =
  \left(
    \frac C T
  \right)^2
  ,
  \label{eq:Lagrangian_const}
\end{equation}
%
for some positive $C$.
%
Integrating the square root, we get
%
\begin{equation}
  \int_{ 0 }^{ q(T) }
    \sqrt{
      \textstyle\sum_{ k = 2 }^n
        \Gamma_k \, \lambda_k^2
        \, u_k^2( q' )
    }
    \;
    d q'
  =
  C \, \frac t T
  ,
  \label{eq:q_opt}
\end{equation}
%
where $c$ is to be determined by
the boundary condition at $t = T$ as
%
\begin{equation}
  C =
  \int_{ 0 }^{ q(t) }
    \sqrt{
      \textstyle\sum_{ k = 2 }^n
        \Gamma_k \, \lambda_k^2
        \, u_k^2( q' )
    }
    \;
    d q'
  .
  \notag
\end{equation}
%
Thus, we have an implicit equation for $q(t)$,
and hence the optimal schedule,
$\alpha(t) = \dot q(t)$.
%
Since we only optimized the asymptotic error, $E_A$,
in the above,
the parameter $q(T)$ can be further adjusted
to minimize the total error, $E = E_R + E_A$.
%
A complete program of computing the optimal schedule
is listed in Appendix \ref{sec:program}.

@@

We can characterize the optimal schedule
geometrically as follows.
%
First, we define the mass function as
$$
  M(Q)
  =
  \sqrt{
    \textstyle\sum_{ k = 2 }^n \Gamma_k \, \lambda_k^2 \, e^{-2 \, \lambda_k \, Q}
  }
  .
$$
and the normalized mass distribution as
\begin{equation}
  m(Q)
  =
  \frac{
    M(Q)
  }
  {
    \int_0^{ q(T) } M(Q') \, d Q'
  }
  =
  \frac{
    M(Q)
  }
  {
    C
  }
  ,
\notag
%\label{eq:mass_distr}
\end{equation}
%
such that
$\int_0^{q(T)} m(Q) \, dQ = 1$.
%
Clearly,
the distribution decreases monotonically with $Q$.

@@

In each class, the updating magnitudes of two schedules
of lengths $T_1$ and $T_2$
at corresponding times $t_1/T_1 = t_2/T_2$
are related by $T_1 \, \alpha_1 = T_2 \, \alpha_2$.
%
Further, by Eqs.
\eqref{eq:error_asym_Lagrangian}
and
\eqref{eq:Lagrangian_const},
we have
%
\begin{equation}
  E_A
  =
  \frac { C^2 } { T }
  =
  \frac 1 T
  \left(
    \int_0^{ q(T) } M(Q) \, dQ
  \right)^2
  ,
\label{eq:error_asym2}
\end{equation}
%
which means the asymptotic error
is inversely proportional to the length
for the same-class schedules,
and we have the scaling relation
$T_1 \, E_{A1} = T_2 \, E_{A2}$.

@@

Many commonly-used updating schemes
can be characterized by a rigid window
centered on the current bin.
%
We call these schemes homogeneous,
or translationally invariant.
%
An example is the Gaussian updating scheme
used in metadynamics.
%
Homogeneous updating schemes have significantly fewer
parameters than the general case,
and the multiplication of the updating matrix,
$\mathbf w$,
can be reduced to a convolution with the window,
or the updating kernel.
%
Thus,
there is a one-to-one correspondence between
the eigenvalues of $\mathbf w$
and the updating kernel.

@@

It is natural to introduce homogeneous updating schemes
for a periodic variable\cite{dama2014}.
%
The updating matrix, $\mathbf w$,
assume the following form
%
\begin{equation}
  w_{ij}
  =
  \mu_{i-j}
  +
  \mu_{i-j+n}
  +
  \mu_{i-j-n}
  ,
\notag
%\label{eq:w_band_pbc}
\end{equation}
%
where
$\mu_{-b}, \dots, \mu_b$ ($b \le n/2$)
are a set of numbers satisfying
%
\begin{equation}
  \mu_{-b} + \cdots + \mu_b = 1
  ,
\label{eq:mu_normalization}
\end{equation}
%
and $\mu_l = 0$ for $|l| > b$.
%
Particularly,
we shall only consider symmetric kernels
that satisfy
%
\begin{equation}
  \mu_i = \mu_{-i}
  .
\label{eq:mu_symm}
\end{equation}
%
If $b = n/2$, then $\mu_{-b}$ is removed
from the sum in Eq. \eqref{eq:mu_normalization}.
%
These numbers constitutes an \emph{updating kernel}.

@@

For the Gaussian updating kernel of width $\sigma$
with $\Delta x \ll \sigma \ll 1$,
we can extend
the upper limit of the integrals
to infinity, and
%
\begin{equation}
  \mu(x)
  =
  \frac{            1            }
       { \sqrt{ 2 \pi \sigma^2 } }
  %
  \exp\left(
        -\frac{       x^2     }
              { 2 \, \sigma^2 }
      \right)
  .
\notag
\end{equation}
%
%
If follows that all eigenvalues are positive:
%
\begin{align}
  \lambda_k
  &=
  \exp\left[
        -\frac{ (k - 1)^2 \, \sigma^2 }
              {           2           }
      \right]
  %\notag
  %\\
  %&=
  %\exp\left[
  %      -\frac{ \pi^2 }{ 2 }
  %      \left(
  %        \frac{ k - 1 }
  %             {   n   }
  %      \right)^2
  %      \left(
  %        \frac{  \sigma }
  %             { \Delta x }
  %      \right)^2
  %    \right]
  .
\notag
%\label{eq:lambda_Gaussian}
\end{align}

@@

Since the first eigenmode corresponds to
a uniform shift of all $v_k$'s,
we can freely set $\lambda_1$ to $\lambda_2$
without changing the nature of the updating scheme.
%
Now an updating matrix
with identical eigenvalues
is a multiple of the identity matrix.
%
Thus,
the single-bin scheme is most efficient
in asymptotic convergence
if no eigenvalue $\lambda_k$ is zero.

@@

As an alternative to the
optimal schedule from Eq. \eqref{eq:q_opt},
here we consider the simpler
inverse-time formula, Eq. \eqref{eq:alpha_invtlambda},
%with a free parameter $\lambda$.
%
whose error can be computed explicitly.

>>>>

The Wang-Landau (WL) algorithm\cite{wang2001, wang2001pre}
and meta\-dy\-nam\-ics\cite{huber1994, laio2002,
laio2008, barducci2011, sutto2012}
are two widely-used and closely-related\cite{micheletti2004}
techniques for this purpose,
in which a bias potential is gradually built up
during simulation.
%
These techniques employ updating loops
to discourage future visits to configurations
with the current $z$ value
by elevating the bias potential there.
%
The main difference
is that in the WL case,
each updating step is limited to the bin
containing the current $z$,
while metadynamics adopts an extended
Gaussian window
covering several neighboring bins.
%
The latter is more often used in MD simulations
as it guarantees a smooth profile
of the bias potential
suitable for numerical differentiation
in deriving the bias force.
%
However, in both cases, the bias potential
is updated frequently (usually every few MC or MD steps),
%which % brings about a side effect of disrupting
and the fluctuation
disrupts \note{``disturbs''?} the underlying
equilibrium sampling dynamics\cite{laio2002}.
%
Thus, one often has to reduce
the updating magnitude
over the course of simulation
to decrease the effect.

The updating schedule,
or the manner of reducing
the updating magnitude,
therefore,
determines the precision of the final bias potential,
hence that of the PMF\cite{liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008, morozov2009,
komura2012, caparica2012, caparica2014,
barducci2008, dickson2011, dama2014}.
%
For the WL algorithm, it is known
that the optimal schedule
is given by the product of the
inverse time\cite{liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008}
(with ``time'' being
the number of simulation steps).
%
The optimal schedule for the metadynamics case,
however, is less studied.

In this study,
we present a method of computing
the optimal schedule
for a general adaptive flat-distribution sampling method.
%
In this method, the optimal schedule is derived
from an equation of motion of a free particle,
with the initial updating magnitude set to
half of the previous equilibrium value.
%
The optimal schedule in the WL case
is shown to be given by
the inverse-time formula,
with the origin of time modified
to satisfy the above optimal initial condition.
%
For metadynamics, however,
the optimal schedule is more complex,
and it can be sensitive to the simulation length
and the width of the Gaussian window.
%
Further, we generalize
the WL updating scheme
to a class of bandpass updating schemes
that are optimal for long simulations,
and the modified inverse-time schedule is optimal for these schemes.
%
%For finite-length simulations, however,
%metadynamics may deliver better performance.

@@

Here we shall call a simulation that targets
a nearly flat distribution\cite{
dayal2004, trebst2004, zhang2007, barducci2008, singh2011},
$p_i$,
a flat-distribution sampling (FDS) simulation.
%
Roughly speaking, there are two types of FDS methods.
%
The first type,
referred to as the equilibrium FDS, % (E-FDS) methods hereinafter,
uses a fixed bias potential, $V_i$,
(which is estimated beforehand)
during simulation.
%
Examples include the original
entropic or multicanonical sampling\cite{
berg1992, lee1993}
and simulated tempering\cite{
marinari1992, lyubartsev1992}
with $i$ being the index of energy
and that of temperature, respectively.
%
The estimated bias potential, $V_i$,
are often improved
by the histogram, $H_i$, accumulated
during simulation.
%
Using $H_i$ as an estimate of $\pi_i$ in
Eq. \eqref{eq:pi_p_phi1}
allows us to correct $V_i$ as
%
\begin{equation}
V^\mathrm{corr}_i
=
V_i
+
\ln \frac{ H_i }
         { p_i }.
\label{eq:vcorr_equil}
\end{equation}
%
\note{This follows from
  $$
  \begin{aligned}
    H_i \approx \pi_i
    &\propto p^*_i \, e^{-V_i},
    \\
    p_i
    &\propto p^*_i \, e^{-V^\mathrm{corr}_i}.
  \end{aligned}
  $$
}
For an unknown system,
it is often difficult
to guess a sufficiently accurate
$V_i$ to start with.
%
Thus,
it can take many iterative rounds of simulations
to achieve the desired flat distribution.
%
\note{Further, a reasonably precise histogram
requires an accumulation period
longer than the autocorrelation time,
which can also be hard to estimate beforehand.}

@@

%For the single-bin scheme
%used in WL algorithm,
Later,
it was found that
a better way
of controlling the schedule $\alpha(t)$
in late stages
is to follow the formula
%
\begin{equation}
  \alpha(t) = \frac{1}{t},
  \label{eq:alpha_invt}
\end{equation}
%
where $t$,
referred to as the ``time'' below,
is the number of steps
from the beginning of the simulation.
%
This simple yet effective formula has attracted
several studies\cite{belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008, morozov2009,
komura2012, caparica2012, caparica2014}.
%
There are, however, some reservation about
the constant of proportionality:
the optimal formula for $\alpha(t)$
might be $c/t$ with a $c$
different from $1.0$\cite{
morozov2007, zhou2008, morozov2009}.
%
Here we will show that the ideal $c$
is $1.0$, but the origin of $t$ should be shifted
to accommodate the initial error.

@@

We now fix the endpoint value $q(T)$
and variate the curve, $q(t)$ ($0 < t < T$).
%
This is equivalent to variating $u$
with fixed endpoints,
$u\bigl( q(0) \bigr)  = e^{- q(T)}$
and
$u\bigl( q(T) \bigr) = 1$.
%
Note also that the first term on the right-hand side
of Eq. \eqref{eq:x2t_average} is fixed during
the process.
%
%We shall further demand that
%%
%\begin{equation}
%  \lim_{t \to \infty} q(t) \to \infty.
%  \label{eq:qt_limit}
%\end{equation}
%%
%Then, for a large $t$,
%the second term on the right-hand side
%of Eq. \eqref{eq:xt_solution} can be neglected, and
%%
%\begin{align}
%\left\langle x^2(t) \right\rangle
%%&=
%%\int_0^t \int_0^t \dot u(t') \, \dot u(t'')
%%    \left\langle \xi(t') \xi(t'') \right\rangle dt'' \, dt'
%%\notag
%%\\
%&=
%\int_0^t \int_0^t
%  \dot u\bigl( q(t') \bigr) \,
%  \dot u\bigl( q(t'') \bigr) \,
%  \kappa(t' - t'') \, dt'' \, dt'.
%\label{eq:x2t_average_asym}
%\end{align}
%
By the Euler-Lagrange equation, we get
$$
\begin{aligned}
0
&=
\frac{d}{dt} \int_0^T
  \dot u\bigl( q(t') \bigr) \, \kappa(t' - t) \, dt'
\\
&= \int_0^T
  \ddot u\bigl( q(t') \bigr) \, \kappa(t' - t) \, dt',
\end{aligned}
$$
where we have dropped the boundary terms
by assuming Eq. \eqref{eq:kappa_limit}
and a large $T$.
%
%In fact, we can repeat the process $n$ times, and
%$$
%\int_0^t u^{(n)}(t') \, \kappa(t' - \tau) \, dt' = 0,
%\qquad (n \ge 1)
%$$
%
In order for this to hold for any $t$,
we must have
%
$
\ddot u\bigl( q(t') \bigr) = 0
,
$
or
\begin{equation}
  \dot u\bigl( q(t') \bigr) = c
  ,
\label{eq:dudt_const}
\end{equation}
%
where $c$ is a constant of $t'$.
%
Using Eq. \eqref{eq:u_def},
we get
$$
e^{ q(t') - q(T) }
=
c \, (t' + t_0),
$$
where $t_0$ is a constant.
%
Taking the logarithm,
and differentiating both sides
with respect to $t'$
yields Eq. \eqref{eq:alpha_invt1}.

@@

%It is worth pointing out
Finally, there is a stability condition.
%
In a \emph{stable} updating scheme,
no eigenmode grows indefinitely,
which means that no eigenvalue $\lambda_k$
can be negative.

@@

In the asymptotic regime
(i.e., if the simulation length, $T$, is sufficiently long),
we can approximate $\kappa_k(t)$ as
%
\begin{equation}
  \kappa_k(t) \approx \Gamma_k \, \delta(t)
  ,
\label{eq:kappa_delta}
\end{equation}
%
where
\begin{equation}
  \Gamma_k = \int_{-\infty}^\infty \kappa_k(t) \, dt,
  \notag
  %\label{eq:Gamma_integral}
\end{equation}
or in discrete time,
\begin{equation}
  \Gamma_k = \sum_{t = -\infty}^\infty \kappa_k(t).
  \label{eq:Gamma_sum}
\end{equation}
%
Then, Eq. \eqref{eq:error_asym} is simplified as
%
\begin{align}
  \Err_A
  &=
  \int_0^T
  \sum_{k = 2}^n
  \Gamma_k \, \dot u_k^2\bigl( q(t) \bigr) \, dt.
\label{eq:error_asym1}
\end{align}
%
In this way,
the underlying sampling process
affects the error only through the numbers,
$\Gamma_k$'s.
%
In Appendix \ref{sec:Gamma},
we evaluate $\Gamma_k$'s in two special cases,
and show how to measure them
from an adaptive FDS simulation.

@@

We shall use this result to model the initial error
before starting the schedule $\alpha(t)$ to be optimized.
%
That is, we assume
$\alpha(t) = \alpha_0$
for $t < 0$, and
the system has entered the equilibrium
under a fixed $\alpha_0$ at $t = 0$.
%
We can then use Eq. \eqref{eq:y2_eql}
for the $\langle y_k^2(0) \rangle$'s
in Eq. \eqref{eq:error_res},
and
the residual error at the end of period $T$
is given by
%
\begin{equation}
  \Err_R
  =
  \sum_{k = 2}^n
      \frac 1 2 \, \alpha_0 \, \Gamma_k \, \lambda_k \,
      e^{  - 2 \, \lambda_k \, q(T) }
  .
  \label{eq:error_res1}
\end{equation}
%
Note that the residual error
depends on $\alpha(t)$ only through
the integral $q(T) = \int_0^T \alpha(t) \, dt$.
%
Thus, if we variate $\alpha(t)$ under a fixed $q(T)$,
this residual part can be ignored,
and the optimization can be done exclusively
on the asymptotic part, $\Err_A$.

@@

As shown above, with a fixed value of $q(T)$,
the optimization of the total error is equivalent to
that of the asymptotic part.
%
We can rewrite Eq. \eqref{eq:error_asym1} as an action
%
\begin{equation}
  \Err_A
  =
  \int_0^T
    {\mathcal L} \bigl[ q(t)\bigr]
    \, dt
  ,
%\notag
\label{eq:error_asym_Lagrangian}
\end{equation}
%
where the Lagrangian,
%
\begin{align}
  {\mathcal L} \bigl[ q(t) \bigr]
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \dot u_k^2\bigl[ q(t) \bigr]
  \notag
  \\
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \lambda_k^2 \, u_k^2 \bigl[ q(t) \bigr]
  \; \dot q^2( t ),
\notag
\end{align}
%
represents a free particle
with a position-dependent mass.
%
Since the Lagrangian
does not explicitly depend on time, $t$,
the Hamiltonian is conserved.
%
But the Hamiltonian
%
\begin{equation}
  \mathcal H
  =
  \frac{ \partial \mathcal L }
       { \partial \dot q     }
  \, \dot q
  -
  \mathcal L
  =
  2 \, \mathcal L
  - \mathcal L
  =
  \mathcal L,
  %=
  %\mathrm{const.}
  \notag
  %\label{eq:error_asym_Hamiltonian}
\end{equation}
%
coincides with the Lagrangian.
%
So we may set
%
\begin{equation}
  \mathcal L
  =
    \sum_{ k = 2 }^n
      \Gamma_k \, \lambda_k^2
      \, u_k^2 \bigl[ q(t) \bigr]
  \;
  \dot q^2(t)
  =
  \left(
    \frac C T
  \right)^2
  ,
  \label{eq:Lagrangian_const}
\end{equation}
%
for some positive $C$,
%
%We may treat Eq. \eqref{eq:Lagrangian_const}
as a generalization of Eq. \eqref{eq:dudt_const}.
%
By integrating the square root, we get
%
\begin{equation}
  \int_{ 0 }^{ q(t) }
    \sqrt{
      \textstyle\sum_{ k = 2 }^n
        \Gamma_k \, \lambda_k^2
        \, u_k^2( q' )
    }
    \;
    d q'
  =
  C \, \frac t T
  ,
  \label{eq:q_opt}
\end{equation}
%
where $C$ is to be determined by
the boundary condition at $t = T$ as
%
\begin{equation}
  C =
  \int_{ 0 }^{ q(T) }
    \sqrt{
      \textstyle\sum_{ k = 2 }^n
        \Gamma_k \, \lambda_k^2
        \, u_k^2( q' )
    }
    \;
    d q'
  .
  \notag
\end{equation}
%
Thus, we have an implicit equation for $q(t)$,
and hence the optimal schedule,
$\alpha(t) = \dot q(t)$.
%
Since we only optimized the asymptotic error, $E_A$,
in the above,
the parameter $q(T)$ can be further adjusted
to minimize the total error, $E = E_R + E_A$
(cf. Sec. \ref{sec:optinitalpha}).
%
A complete program of computing the optimal schedule
is listed in Appendix \ref{sec:program}.

@@

We can characterize the optimal schedule
geometrically as follows.
%
First, we define the mass function as
$$
  M(Q)
  =
  \sqrt{
    \textstyle\sum_{ k = 2 }^n \Gamma_k \, \lambda_k^2 \, e^{-2 \, \lambda_k \, Q}
  }
  .
$$
and the normalized mass distribution as
\begin{equation}
  m(Q)
  =
  \frac{
    M(Q)
  }
  {
    \int_0^{ q(T) } M(Q') \, d Q'
  }
  =
  \frac{
    M(Q)
  }
  {
    C
  }
  ,
\notag
%\label{eq:mass_distr}
\end{equation}
%
such that
$\int_0^{q(T)} m(Q) \, dQ = 1$.
%
Clearly,
both $M(Q)$ and $m(Q)$ decrease monotonically with $Q$.

@@

In each class, the updating magnitudes of two schedules
of lengths $T_1$ and $T_2$
at corresponding times $t_1/T_1 = t_2/T_2$
are related by $T_1 \, \alpha_1 = T_2 \, \alpha_2$.
%
Further, by Eqs.
\eqref{eq:error_asym_Lagrangian}
and
\eqref{eq:Lagrangian_const},
we have
%
\begin{equation}
  E_A
  =
  \frac { C^2 } { T }
  =
  \frac 1 T
  \left(
    \int_0^{ q(T) } M(Q) \, dQ
  \right)^2
  ,
\label{eq:error_asym2}
\end{equation}
%
which means the asymptotic error
is inversely proportional to the simulation length
for the same-class schedules,
i.e., we have the scaling relation
$T_1 \, E_{A1} = T_2 \, E_{A2}$.

@@

Many commonly-used updating schemes
can be characterized by a rigid window
centered on the current bin.
%
We call these schemes homogeneous,
or translationally invariant.
%
An example is the Gaussian updating scheme
used in metadynamics.
%
Homogeneous updating schemes
share the same set of eigenvectors
but not the corresponding eigenvalues.
%
They have significantly fewer
parameters than the general case,
and the multiplication of the updating matrix,
$\mathbf w$,
can be reduced to a convolution with the window,
or the updating kernel.
%
Thus,
there is a one-to-one correspondence between
the eigenvalues of $\mathbf w$
and the updating kernel.

@@

It is natural to introduce homogeneous updating schemes
for a periodic variable\cite{dama2014}.
%
The updating matrix, $\mathbf w$,
assume the following form
%
\begin{equation}
  w_{ij}
  =
  \mu_{i-j}
  +
  \mu_{i-j+n}
  +
  \mu_{i-j-n}
  ,
\notag
%\label{eq:w_band_pbc}
\end{equation}
%
where
$\mu_{-b}, \dots, \mu_b$ ($b \le n/2$)
are a set of numbers satisfying
%
\begin{equation}
  \mu_{-b} + \cdots + \mu_b = 1
  ,
\label{eq:mu_normalization}
\end{equation}
%
and $\mu_l = 0$ for $|l| > b$.
%
Particularly,
we shall only consider symmetric kernels
that satisfy
%
\begin{equation}
  \mu_i = \mu_{-i}
  .
\label{eq:mu_symm}
\end{equation}
%
If $b = n/2$, then $\mu_{-b}$ is removed
from the sum in Eq. \eqref{eq:mu_normalization}
to avoid double counting.
%
These numbers constitutes an \emph{updating kernel}.

@@

For the Gaussian updating kernel,
if the width $\sigma$ satisifies
with $\Delta x \ll \sigma \ll 1$,
we can extend
the upper limit of the integrals
to infinity, and
%
\begin{equation}
  \mu(x)
  =
  \frac{            1            }
       { \sqrt{ 2 \pi \sigma^2 } }
  %
  \exp\left(
        -\frac{       x^2     }
              { 2 \, \sigma^2 }
      \right)
  .
\notag
\end{equation}
%
%
If follows that all eigenvalues are positive:
%
\begin{align}
  \lambda_k
  &=
  \exp\left[
        -\frac{ (k - 1)^2 \, \sigma^2 }
              {           2           }
      \right]
  %\notag
  %\\
  %&=
  %\exp\left[
  %      -\frac{ \pi^2 }{ 2 }
  %      \left(
  %        \frac{ k - 1 }
  %             {   n   }
  %      \right)^2
  %      \left(
  %        \frac{  \sigma }
  %             { \Delta x }
  %      \right)^2
  %    \right]
  .
\notag
%\label{eq:lambda_Gaussian}
\end{align}

@@

Since the first eigenmode corresponds to
a uniform shift of all $v_k$'s,
we can freely set $\lambda_1$ to $\lambda_2$
without changing the nature of the updating scheme.
%
Now an updating matrix
with identical eigenvalues
is a multiple of the identity matrix.
%
Thus,
the single-bin scheme is most efficient
in asymptotic convergence.
%if no eigenvalue $\lambda_k$ is zero.

@@

As an alternative to the
optimal schedule from Eq. \eqref{eq:q_opt},
here we consider the modified
inverse-time formula, Eq. \eqref{eq:alpha_invtlambda},
%with a free parameter $\lambda$.
%
whose error can be computed explicitly.
